\section{Conclusion}\label{sec:conclusion}

We have developed a residual-based \emph{a posteriori} error estimator for
physics-informed neural network (PINN) approximations of coercive elliptic
partial differential equations in the energy seminorm. The key contribution
of this work is the systematic incorporation of boundary mismatch through
a harmonic lifting term, leading to a \emph{reliable and fully computable}
estimator that remains consistent with the PINN formulation based on soft
Dirichlet enforcement.

The proposed estimator is given by
\[
\|\nabla(u - \hat{u})\|_{L^2(\Omega)}
\;\le\;
\frac{1}{\alpha}\|R\|_{H^{-1}(\Omega)}
+
\|\nabla w\|_{L^2(\Omega)},
\]
where the dual norm of the residual captures interior physics violation,
and the lifting term quantifies boundary mismatch propagated into the domain.
This decomposition reveals a fundamental limitation of standard PINN training:
\emph{minimization of the training loss does not guarantee control of the
true approximation error in Sobolev norms}.

Theoretical analysis establishes reliability of the estimator for
constant-coefficient problems. For variable-coefficient diffusion,
a Laplace-based harmonic lifting is used as a computable surrogate,
resulting in a practically reliable upper bound consistent with the
implemented numerical pipeline.

A comprehensive and fully reproducible numerical study confirms both
reliability and sharpness of the estimator. Across all benchmarks
(Table~\ref{tab:benchmark_results}), the effectivity index satisfies
\[
1 \le \eta \le 1.31,
\]
demonstrating near-optimal performance. In particular, the ablation study
(Table~\ref{tab:ablation_results}) shows that the residual-only estimator
severely underestimates the true error, with effectivity indices as low as
$\eta_r \approx 0.0275$ for smooth problems. In contrast, the inclusion of
the boundary lifting term restores reliability and yields accurate error
certification.

A central finding of this work is that \emph{boundary mismatch is the dominant
source of error in PINNs with soft Dirichlet enforcement}. In all considered
benchmarks, the lifting contribution exceeds the residual contribution by
one to two orders of magnitude, providing a quantitative explanation for the
failure of residual-based indicators and training loss as measures of
solution accuracy.

From a computational perspective, the estimator is fully realizable using
efficient numerical solvers, including discrete sine transform (DST)
methods on structured domains and finite element discretizations on
irregular geometries. The entire pipeline is reproducible and relies only
on the trained PINN solution and problem data, without requiring access to
the exact solution for estimator evaluation.

This work establishes a principled and practically implementable framework
for \emph{certifying} PINN solutions. By integrating classical ideas from
a posteriori error estimation with modern neural PDE solvers, it resolves
a critical gap between scientific machine learning and numerical analysis.

\medskip

\noindent
\textbf{Future work.}
Several important directions arise from this study:
\begin{itemize}
\item Development of estimators based on $a$-harmonic lifting to achieve
full theoretical consistency for variable-coefficient problems.
\item Design of adaptive sampling and training strategies guided by the
estimator for improved accuracy and efficiency.
\item Extension to time-dependent, nonlinear, and convection-dominated
problems, where PINNs exhibit significant stability challenges.
\item Integration of the estimator into training loops for
\emph{error-controlled} and certified PINN solvers.
\end{itemize}

\medskip

\noindent
In conclusion, reliable error certification for PINNs requires explicit
accounting of both interior residuals and boundary mismatch. The estimator
proposed in this work provides a mathematically rigorous, computationally
efficient, and experimentally validated tool for achieving this goal,
thereby advancing the development of trustworthy neural solvers for
partial differential equations.