\section{Computable Residual-Based Estimator}\label{sec:estimator}

This section provides the \emph{computational realization} of the estimator
derived in Section~\ref{sec:analysis}. We describe how each component is
evaluated in practice in a manner consistent with the implementation.

\subsection{Estimator definition}

Recall that the a posteriori error estimator is given by
\begin{equation}\label{eq:estimator-final}
\eta :=
\frac{1}{\alpha}\|R\|_{H^{-1}(\Omega)}
+
\|\nabla w\|_{L^2(\Omega)},
\end{equation}
where
\[
R = f + \nabla \cdot (a \nabla \hat{u})
\]
and $w$ is the harmonic lifting defined in
\eqref{eq:analysis-lifting}.

The estimator consists of two fully computable components:
(i) a dual norm of the interior residual, and
(ii) an energy norm of the boundary lifting.

\subsection{Computation of the dual norm}

The dual norm $\|R\|_{H^{-1}(\Omega)}$ is computed via the Riesz
representation. Specifically, we solve the auxiliary problem:
\begin{equation}\label{eq:dual-problem}
\begin{aligned}
- \Delta z &= R && \text{in } \Omega,\\
z &= 0 && \text{on } \partial\Omega.
\end{aligned}
\end{equation}

Then,
\begin{equation}\label{eq:dual-norm}
\|R\|_{H^{-1}(\Omega)} = \|\nabla z\|_{L^2(\Omega)}.
\end{equation}

\paragraph{Numerical realization.}

\begin{itemize}
\item \textbf{Structured domains:}
For $\Omega = (0,1)^2$, we discretize \eqref{eq:dual-problem}
on a uniform grid and solve using the Discrete Sine Transform (DST),
yielding an efficient $\mathcal{O}(N \log N)$ solver.

\item \textbf{Irregular domains:}
For the L-shaped domain, we employ a conforming finite element
discretization with piecewise linear basis functions.

\end{itemize}

The residual $R$ is evaluated pointwise using automatic differentiation
of the neural network $\hat{u}$.

\subsection{Computation of the boundary lifting}

The lifting $w$ is obtained by solving
\begin{equation}\label{eq:lifting-compute}
\begin{aligned}
- \Delta w &= 0 && \text{in } \Omega,\\
w &= g - \hat{u} && \text{on } \partial\Omega.
\end{aligned}
\end{equation}

\paragraph{Numerical realization.}

\begin{itemize}
\item On structured grids, the harmonic extension is computed using
spectral methods consistent with the DST framework.

\item On unstructured domains, the lifting is computed using finite
elements with Dirichlet boundary data imposed strongly.
\end{itemize}

The energy contribution is then computed as
\begin{equation}
\|\nabla w\|_{L^2(\Omega)}.
\end{equation}

\subsection{Algorithmic pipeline}

The complete estimator evaluation proceeds as follows:

\begin{enumerate}[leftmargin=*]

\item Evaluate $\hat{u}$ and its derivatives using automatic differentiation.

\item Compute the residual field
\[
R = f + \nabla \cdot (a \nabla \hat{u}).
\]

\item Solve the dual problem \eqref{eq:dual-problem} to obtain $z$.

\item Compute the dual norm:
\[
\|R\|_{H^{-1}} = \|\nabla z\|.
\]

\item Solve the lifting problem \eqref{eq:lifting-compute} to obtain $w$.

\item Compute the lifting contribution:
\[
\|\nabla w\|.
\]

\item Assemble the estimator:
\[
\eta = \frac{1}{\alpha}\|R\|_{H^{-1}} + \|\nabla w\|.
\]

\end{enumerate}

\subsection{Consistency with implementation}

The above procedure is implemented in the accompanying code repository.
In particular:

\begin{itemize}
\item Automatic differentiation is used for exact evaluation of
$\nabla \hat{u}$ and $\nabla \cdot (a \nabla \hat{u})$.

\item The dual problem and lifting problem are solved using
numerically consistent discretizations (DST or FEM depending on geometry).

\item No access to the exact solution $u$ is required.
\end{itemize}

\subsection{Relation to experimental results}

The estimator defined in \eqref{eq:estimator-final} is used to compute
the values reported in Table~\ref{tab:benchmark_results} and
Table~\ref{tab:ablation_results}. :contentReference[oaicite:0]{index=0}

The decomposition into residual and lifting contributions clearly shows that:

\begin{itemize}
\item The residual-only estimator severely underestimates the error,
\item The lifting term dominates the total error,
\item The full estimator achieves effectivity indices close to one.
\end{itemize}

\subsection{Effectivity index}

We define the effectivity index as
\begin{equation}
\eta_{\mathrm{eff}} :=
\frac{\eta}{\|\nabla(u - \hat{u})\|_{L^2(\Omega)}}.
\end{equation}

Across all benchmarks, we observe
\[
1 \le \eta_{\mathrm{eff}} \le 1.31,
\]
demonstrating that the estimator is both reliable and sharp.